Differentiable Optics with {\Large $\partial$}Lux: I - Deep Calibration of Flat Field and Phase Retrieval with Automatic Differentiation

### 2023-04-14 src/tex/ms.tex
[showyourwork] overleaf sync

\newcommand\jax{\textsc{Jax}\xspace}
\newcommand\dlux{$\partial$Lux\xspace}
\newcommand\equinox{\texttt{equinox}\xspace}
\newcommand\optax{\texttt{optax}\xspace}
\newcommand\numpyro{\texttt{numpyro}\xspace}
\newcommand\kepler{\textit{Kepler}\xspace}

@@

\begin{abstract}
% phase retrieval and detector calibration are important limiting factors in many space missions
The sensitivity limits of space telescopes are imposed by uncalibrated errors in the point spread function, photon-noise, background light, and detector sensitivity. 
%these are hard to calibrate
These are typically calibrated with specialized wavefront sensor hardware and with flat fields obtained on the ground or with calibration sources, but these leave vulnerabilities to residual time-varying or non-common path aberrations and variations in the detector conditions. It is therefore desirable to infer these from science data alone, facing the prohibitively high dimensional problems of phase retrieval and pixel-level calibration.
% we introduce a new package dLux for differentiable, gpu-accelerated physical optics simulation
We introduce a new Python package for physical optics simulation, \dlux, which uses the machine learning framework \jax to achieve GPU acceleration and automatic differentiation, and apply this to simulating astronomical imaging.
% we show we can do flat field and phase retrieval simultaneously
In this first of a series of papers, we show that gradient descent enabled by autodiff can be used to simultaneously perform phase retrieval and calibration of detector sensitivity, scaling efficiently to inferring millions of parameters.
% this is the first of a series of applications
This new framework enables high dimensional optimization and inference in data analysis and hardware design in astronomy and beyond, which we explore in subsequent papers in this series.
\end{abstract}

@@

In this paper we introduce a new optical simulation package, \dlux, available under an open-source BSD 3 license. It significantly extends the capabilities of \texttt{morphine} \cite{pope2021,phase_ret_and_design}, a previous \jax optics package from our team based on \texttt{poppy}, being rewritten from the ground up taking full advantage of more-recent \jax features and libraries.  It is built in an object-oriented \jax framework called \equinox \cite{kidger2021equinox}, with an optical system consisting of an \equinox module wrapping a stack of `layers' which are themselves \equinox modules applying phase or amplitude screens, or performing Fourier or Fresnel transforms on a wavefront. This allows us to re-frame the idea of a computational optical model by analogy to a `parametrized neural network', with each operation performed on the wavefront analogous to a neural network layer encoding the physics of the transformation. The object-oriented nature of \equinox allows for larger and more complex models to be build and seamlessly integrated than would be possible within the purely functional framework default to \jax. \equinox is also built to be directly integrated into the \jax framework, allowing for all \jax function to seamlessly interface with any \equinox models and vice-versa. Talk about zodiax too? We use the optimization library \optax \cite{optax2020github} for gradient-descent optimization, and the probabilistic programming library \numpyro \cite{Phan2019} for \ac{hmc}.

@@

To recover these parameters we initialise a naive unaberrated optical model with a uniform \ac{prf}. Stellar positions were perturbed by adding a random value drawn from a normal distribution mean 0 deviation 1 pixel and fluxes were perturbed by multiplying their value by a random value value drawn from a normal distribution mean 1 deviation 0.1. We will use this model to recover the true values. Figure~\ref{fig:data_residual} shows the data and the initial and final residuals.

@@

\begin{figure*}
    \centering
    \includegraphics[width=1.0\textwidth]{figs/noise_performance.pdf}
    % \includegraphics[width=0.8\textwidth]{figures/noise_performance.pdf}
    \caption{Recovered parameter error relative to photon noise expectation. Left: Mean relative per-mode Zernike error. Middle: Mean relative flux error. Right: Mean relative positional error.}
    \label{fig:noise}
\end{figure*}

@@

It is important to note in this section that there is a degree of scatter in the recovered results. Due to the gradient descent optimisation strategy there is a degree of sensitivity to the optimisation hyper-parameters, such as learning rates which must be tuned based on the flux, momentum, and distance to the noise floor. These parameters are simple enough to tune for individual data sets with similar scales, but it is non-trivial to automate this process for data with a vast range of scales like those being analysed here. The hyper-parameters chosen were tuned on the 1e8 flux data set and roughly scaled to the work with the other data sets, which is why we observe the best recovery for that set of parameters. These figures are designed to demonstrate the overall trends in the different noise regimes, rather than to be a precise analysis of the fundamental limits.

>>>>

\newcommand\jax{\textsc{Jax}\xspace}
\newcommand\dlux{$\partial$Lux\xspace}
\newcommand\equinox{\texttt{equinox}\xspace}
\newcommand\zodiax{\texttt{zodiax}\xspace}
\newcommand\optax{\texttt{optax}\xspace}
\newcommand\numpyro{\texttt{numpyro}\xspace}
\newcommand\kepler{\textit{Kepler}\xspace}

@@

\begin{abstract}
% phase retrieval and detector calibration are important limiting factors in many space missions
The sensitivity limits of space telescopes are imposed by uncalibrated errors in the point spread function, photon-noise, background light, and detector sensitivity. 
%these are hard to calibrate
These are typically calibrated with specialized wavefront sensor hardware and with flat fields obtained on the ground or with calibration sources, but these leave vulnerabilities to residual time-varying or non-common path aberrations and variations in the detector conditions. It is therefore desirable to infer these from science data alone, facing the prohibitively high dimensional problems of phase retrieval and pixel-level calibration.
% we introduce a new package dLux for differentiable, gpu-accelerated physical optics simulation
We introduce a new Python package for physical optics simulation, \dlux, which uses the machine learning framework \jax to achieve GPU acceleration and automatic differentiation ('autodiff'), and apply this to simulating astronomical imaging.
% we show we can do flat field and phase retrieval simultaneously
In this first of a series of papers, we show that gradient descent enabled by autodiff can be used to simultaneously perform phase retrieval and calibration of detector sensitivity, scaling efficiently to inferring millions of parameters.
% this is the first of a series of applications
This new framework enables high dimensional optimization and inference in data analysis and hardware design in astronomy and beyond, which we explore in subsequent papers in this series.
\end{abstract}

@@

In this paper we introduce a new optical simulation package, \dlux, available under an open-source BSD 3 license. It significantly extends the capabilities of \texttt{morphine} \cite{pope2021,phase_ret_and_design}, a previous \jax optics package from our team based on \texttt{poppy}, being rewritten from the ground up taking full advantage of more-recent \jax features and libraries.  It is built in an object-oriented \jax framework called \zodiax which extends \equinox \cite{kidger2021equinox} for scientific programming. An optical system consisting of an \zodiax module wrapping a stack of `layers' which are themselves \zodiax modules applying phase or amplitude screens, or performing Fourier or Fresnel transforms on a wavefront. This allows us to re-frame the idea of a computational optical model by analogy to a `parametrized neural network', with each operation performed on the wavefront analogous to a neural network layer encoding the physics of the transformation. The object-oriented nature of \zodiax allows for larger and more complex models to be built and seamlessly integrated than would be possible within the purely functional framework default to \jax. \zodiax is also built to be directly integrated into the \jax framework, allowing for all \jax functions to seamlessly interface with any \zodiax models and vice-versa. Talk about zodiax too? We use the optimization library \optax \cite{optax2020github} for gradient-descent optimization, and the probabilistic programming library \numpyro \cite{Phan2019} for \ac{hmc}.

@@

To recover these parameters we initialise a naive unaberrated optical model with a uniform \ac{prf}. Stellar positions were perturbed by adding a random value drawn from a normal distribution mean 0 deviation 1 pixel and fluxes were perturbed by multiplying their value by a random value drawn from a normal distribution mean 1 deviation 0.1. We will use this model to recover the true values. Figure~\ref{fig:data_residual} shows the data and the initial and final residuals.

@@

\begin{figure*}
    \centering
    \includegraphics[width=1.0\textwidth]{figures/noise_performance.pdf}
    % \includegraphics[width=0.8\textwidth]{figures/noise_performance.pdf}
    \caption{Recovered parameter error relative to photon noise expectation. Left: Mean relative per-mode Zernike error. Middle: Mean relative flux error. Right: Mean relative positional error.}
    \label{fig:noise}
\end{figure*}

@@

It is important to note in this section that there is a degree of scatter in the recovered results. Due to the gradient descent optimisation strategy there is a degree of sensitivity to the optimisation hyper-parameters, such as learning rates which must be tuned based on the flux, momentum, and distance to the noise floor. These parameters are simple enough to tune for individual data sets with similar scales, but it is non-trivial to automate this process for data with a vast range of scales like those being analysed here. The hyper-parameters chosen were tuned on the 1e8 flux data set and roughly scaled to work with the other data sets, which is why we observe the best recovery for that set of parameters. These figures are designed to demonstrate the overall trends in the different noise regimes, rather than to be a precise analysis of the fundamental limits.
